Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2014-05-26 paper.tex
added minor

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  
This appears to be the first time that the statistical properties of a non data-aided estimator for carrier amplitude have been analysed theoretically.
%Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  
The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK), and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e., $P \cap D = \emptyset$  where $\emptyset$ is the empty set. We let $L = \abs{P \cup D}$ be the total number of symbols transmitted.

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$.  By definition $\rho_0$ is a positive real number.  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver and must also be estimated.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990,580211,974597,Mengali_andre_synchro_book}.  A practical approach is the least squares estimator, that is, the minimiser of the sum of squares function
\begin{equation}\label{eq:LSdefn}
\begin{split}
LS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

% The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non data-aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
% \begin{equation}\label{eq:viterbiviterbi}
% C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
% \end{equation}
% Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
% This problem does not occur with the least square estimator.

@@

Aside from this extention of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:LSdefn}.  In the literature, analytic study of non-data-aided estimators of carrier phase has primarily involved the derivation of Cram\'{e}r-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,950345,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cram\'{e}r-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptoticially as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

For the case of carrier amplitude the situations is even more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that known non-data-aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Neverthess, Cram\'{e}r-Rao bound have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non-data-aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convinving response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as mulitple anntannae systems, systems employing orthogonal frequency division multiplexing, and systems that must widthstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

That this paper contains novel theoretical developments does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares estimator can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing system parameters, such as the number and placement of pilot symbols, and also provides useful information to other reciever components, such as channel tracking systems.

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal that contains both pilot signals, known to the receiver, and data signals, unknown to the receiver.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  
This appears to be the first time that the statistical properties of a non-data-aided estimator for carrier amplitude have been analysed theoretically.
%Our theoretical results can also be applied to the case where no pilot symbols exist, i.e., noncoherent detection.  
The results of Monte Carlo simulations are provided and these agree with the theoretical results.   
\end{abstract}
\begin{IEEEkeywords}
Detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled and we focus on estimating the phase shift and attenuation.  Methods for estimating the time offset are given in~\cite{Massey1972optimumframe,Oerder_synch_square_circstat_2008,McKilliam_time_offset_pilots_data_2013}. The transmitted symbols are assumed to come from an $M$-ary phase shift keying ($M$-PSK) constellation.  That is, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 2$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e., $P \cap D = \emptyset$  where $\emptyset$ is the empty set. We let $L = \abs{P \cup D}$ be the total number of symbols transmitted.

We assume that %time offset estimation has been performed and that 
$L$ noisy $M$-PSK symbols are observed by the receiver,  %The received signal after matched filtering is,
\begin{equation}\label{eq:sigmod}
y_i = a_0 s_i + w_i, \qquad i \in P \cup D,
\end{equation}
where $w_i$ is noise and $a_0 = \rho_0 e^{j\theta_0}$ is a complex number representing both carrier phase $\theta_0$ and amplitude $\rho_0$.  By definition $\rho_0$ is a positive real number.  Our aim is to estimate $a_0$ from the noisy symbols $\{ y_i, i \in P \cup D \}$.  Complicating matters is that the data symbols $\{d_i, i \in D\}$ are not known to the receiver.  Estimation problems of this type have undergone extensive prior study~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990,580211,974597,Mengali_andre_synchro_book}.  A practical approach is the least squares estimator, that is, the minimiser of the sum of squares function
\begin{equation}\label{eq:LSdefn}
\begin{split}
LS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2  \\
&= \sum_{i \in P} \abs{ y_i - a p_i }^2 + \sum_{i \in D} \abs{ y_i - a d_i }^2,
\end{split}
\end{equation}
where $\abs{x}$ denotes the magnitude of the complex number $x$.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  However, as we show, the estimator works well under less stringent assumptions.  %It is the least squares estimator that we primarily study in this paper.

% The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} (also know as \emph{blind}, \emph{non-data-aided} or \emph{non-decision directed detection}) where no pilot symbols exist ($P = \emptyset$).  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is estimator of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the received signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{C}$ where $\angle$ denotes the complex argument and
% \begin{equation}\label{eq:viterbiviterbi}
% C = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
% \end{equation}
% Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat of this estimator is that it is not obvious how pilot symbols should be included. %Techniques are available but they are somewhat ad hoc~\cite{Cowley_ref_sym_carr_1998}.  
% This problem does not occur with the least square estimator.

@@

Aside from this extension of Mackenthun's algorithm, the primary purpose of this paper is to analyse the least squares estimators of carrier phase and amplitude obtained by minimising~\eqref{eq:LSdefn}.  In the literature, analytic study of non-data-aided estimators of carrier phase has primarily involved the derivation of Cram\'{e}r-Rao lower bounds under the assumption that the noise $\{w_i\}$ is complex Gaussian distributed~\cite{Delmas_exact_crb_psk_2008,950345,Steendam_lowsnr_crb_2001,Cowley_crbs_phase_freq_1996,DAndrea_modified_bounds_1994}.  The bounds themselves are often complicated to compute and analyse and for this reason it is common to make approximations, known as \emph{modified} Cram\'{e}r-Rao bounds~\cite{DAndrea_modified_bounds_1994,Pollok_mckilliam_crb_cpm_2013}, or to analyse bounds that are valid asymptotically as the noise variance increases to infinity or decreases to zero~\cite{Moeneclaey_limit_crbs_1998,Steendam_lowsnr_crb_2001,Delmas_exact_crb_psk_2008}.  %To the authors' knowledge, all of the existing bounds are applicable only in the fully non-data-aided case or in the completely data aided case, that is, when either no pilot symbols exist, or when no data symbols exist.  Deriving bounds for the common practical scenario where there is a mixture of pilot and data symbols might be possible, but would further complicate already complicated bounds.

For the case of carrier amplitude the situation is more complicated.  A simulation study conducted by Pauluzzi~\cite{Pauluzzi2000} suggests that known non-data-aided estimators for carrier amplitude exhibit bias.  The bias is often small when the noise variance is small, but typically becomes significant as the noise variance increases.  Neverthess, Cram\'{e}r-Rao bounds have been derived for \emph{unbiased} estimators of carrier amplitude~\cite{Alagha_crb_snr_bpsk_qpsk_2001,Delmas_exact_crb_psk_2008}.  Unfortunately, these bounds may not be meaningful since known non-data-aided amplitude estimators appear to be biased~\cite{Pauluzzi2000}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cram\'{e}r-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convincing response is that, while the specific results given here apply only to the least squares estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010}, would potentially allow this.  
More ambitiously, our techniques might allow analysis of non-data-aided channel estimators for modern communications systems such as mulitple antenna systems, systems employing orthogonal frequency division multiplexing, and systems that must withstand rapidly time varying channels~\cite{4138046,inthft:204506}.  We see the results presented here as a starting point for a potentially potent approach to the analysis of channel estimators that is complementary to the commonplace method of comparing results of Monte-Carlo simulations with Cram\'{e}r-Rao lower bounds.

That this paper contains novel theoretical developments does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided the design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares estimator can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing system parameters, such as the number and placement of pilot symbols, and also provides useful information to other receiver components, such as channel tracking systems.

### 2014-05-26 response2.tex
added minor

\usepackage[pdftex]{graphicx}
%\usepackage{thumbpdf}
%\usepackage[naturalnames]{hyperref}
\usepackage{xr}
\externaldocument{../async}

@@

\item \textbf{Comment:} Consistently write ``non-data-aided'' throughout the whole text \\
\textbf{Response:}

\item \textbf{Comment:} Page (P) 42, right column (RC), line (L) 50: Aside from this extension of \dots \\
\textbf{Response:}

\item \textbf{Comment:} P 43, LC, L 44: \dots convincing response \dots \\
\textbf{Response:}

\item \textbf{Comment:} P 43, RC, L 2+4: \dots as multiple antenna systems \dots must withstand rapidly \dots \\
\textbf{Response:}

\item \textbf{Comment:} P 43, RC, L 21+28: \dots aided the design of \dots to other receiver components \dots \\
\textbf{Response:}

@@

\item \textbf{Comment:} In the abstract: non data-aided $\to$ change to: non-data-aided \\
\textbf{Response:}

\item \textbf{Comment:} ``we assume that the time-offset has been previously handled'' $\to$ perhaps some motivation for this assumption and the consequences with respect to the system model would be in place \\
\textbf{Response:}

\item \textbf{Comment:} ``such as binary phase shift keying (BPSK), \dots'' $\to$ mentioning BPSK and QPSK apart from general M-PSK is pure overhead in this stage, I would drop it \\
\textbf{Response:}

\item \textbf{Comment:} ``the data symbols are not known to the receiver and must also be estimated'' $\to$ I would drop the second part of the sentence because (e.g. Viterbi and Vterbi estimator) non-data-aided estimators typically do not actually estimate the unknown data symbols \\
\textbf{Response:}

\item \textbf{Comment:} p. 3: ``bounds that are valid asymptoticially'': typo in asymptotically \\
\textbf{Response:}

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data aided case \dots'' --> bounds for the common scenario where there is a mixture of pilot and data symbols can be found in "N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005. \\
\textbf{Response:}

\item \textbf{Comment:} p. 3: ``the situations is'' --> drop `s' at the end of situations \\
\textbf{Response:}

@@

\item \textbf{Comment:} p. 3: last line: typo in ``convincing'' \\
\textbf{Response:}

\item \textbf{Comment:} p. 4: ``systems such as mulitple anntannae'' $\to$ should be ``multiple antennae'' \\
\textbf{Response:}

\item \textbf{Comment:} p. 4: ``the other reciever components'' $\to$ should be ``receiver components'' \\
\textbf{Response:}

@@

\bibliography{../../../bib/bib}

>>>>

\usepackage[pdftex]{graphicx}
%\usepackage{thumbpdf}
%\usepackage[naturalnames]{hyperref}
\usepackage{xr}
\externaldocument{paper}

@@

\item \textbf{Comment:} Consistently write ``non-data-aided'' throughout the whole text \\
\textbf{Response:} Done.

\item \textbf{Comment:} Page (P) 42, right column (RC), line (L) 50: Aside from this extension of \dots \\
\textbf{Response:} Fixed.

\item \textbf{Comment:} P 43, LC, L 44: \dots convincing response \dots \\
\textbf{Response:} Fixed.

\item \label{eq:rev1:multantenna} \textbf{Comment:} P 43, RC, L 2+4: \dots as multiple antenna systems \dots must withstand rapidly \dots \\
\textbf{Response:} Fixed.

\item \textbf{Comment:} P 43, RC, L 21+28: \dots aided the design of \dots to other receiver components \dots \\
\textbf{Response:} Fixed.

@@

\item \textbf{Comment:} In the abstract: non data-aided $\to$ change to: non-data-aided \\
\textbf{Response:} Fixed.

\item \textbf{Comment:} ``we assume that the time-offset has been previously handled'' $\to$ perhaps some motivation for this assumption and the consequences with respect to the system model would be in place \\
\textbf{Response:}  There exist numerous estimators for the time offset in the literature.  We have added the sentence

``Methods for estimating the time offset are given in~\cite{Massey1972optimumframe,Oerder_synch_square_circstat_2008,McKilliam_time_offset_pilots_data_2013}.''

to inform the reader of some of these methods.  The interested reader will find further discussion about time offset estimation within these references.  The assumption of an accurate time offset estimator is common in this literature~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.

\item \textbf{Comment:} ``such as binary phase shift keying (BPSK), \dots'' $\to$ mentioning BPSK and QPSK apart from general M-PSK is pure overhead in this stage, I would drop it \\
\textbf{Response:} Done.

\item \textbf{Comment:} ``the data symbols are not known to the receiver and must also be estimated'' $\to$ I would drop the second part of the sentence because (e.g. Viterbi and Vterbi estimator) non-data-aided estimators typically do not actually estimate the unknown data symbols \\
\textbf{Response:} Done.

\item \textbf{Comment:} p. 3: ``bounds that are valid asymptoticially'': typo in asymptotically \\
\textbf{Response:} Fixed.

\item \textbf{Comment:} p. 3: ``all of the exiting bounds are applicable only in the fully non-data-aided case or in the completely data aided case \dots'' --> bounds for the common scenario where there is a mixture of pilot and data symbols can be found in "N. Noels, H. Steendam, M. Moeneclaey en H. Bruneel, Carrier Phase and Frequency Estimation for Pilot Symbol Assisted Transmission: Bounds and Algorithms, IEEE TSP, Vol. 53, No. 12, p. 4578-4587, Dec. 2005. \\
\textbf{Response:} BLERG. I've removed this paragraph from the introduction but will have to update the code and simulations.

\item \textbf{Comment:} p. 3: ``the situations is'' --> drop `s' at the end of situations \\
\textbf{Response:} Fixed.

@@

\item \textbf{Comment:} p. 3: last line: typo in ``convincing'' \\
\textbf{Response:} Fixed.

\item \textbf{Comment:} p. 4: ``systems such as mulitple anntannae'' $\to$ should be ``multiple antennae'' \\
\textbf{Response:} Reviewer 1 comment~\eqref{eq:rev1:multantenna} wanted ``antenna'' and I have used this spelling.

\item \textbf{Comment:} p. 4: ``the other reciever components'' $\to$ should be ``receiver components'' \\
\textbf{Response:} Fixed.

@@

\input{paper.bbl}
